---
title: "PittsburghBridges"
format: html
---

Approach:

I'll start by creating a beachhead, where I pull down the CSV file from a folder in a public
repo that I created, and dump the contents out using the gt() command.

I'll include an explanation for why I chose this dataset--I grew up in a Pittsburgh river town, where you couldn't get out of town by crossing a bridge.

For code, I'll use Garret Grolemund's book to create a basic program that fulfills the requirement deliverables.
I'll then go through a series of interactions with Claude and/or ChatGPT to:
(1) COnvert the code to tidyverse
(2) EDA.  Look at different choices for river.  Use my domain knowledge to make a decision about the included Yough river.
(3) Feature engineering.  Bin time up into before 1900 and 1900 and after.
(2) Ask and answer a question about Wooden Bridges vs. Metal Bridges over time.
(3) If time allows, I'll attempt to augment the data by getting latitudes and longitudes for each bridge and showing on a google map.
(4) It would be very nice to show a visual where the bridges are dynamically overlaid over time.

```{r}
# install.packages("gt")
library(tidyverse)
library(gt)

```

Beachhead - verify that I can read data into a tibble (dataframe), and dump out contents
Note: I'm using the raw version of https://raw.githubusercontent.com/acatlin/data/refs/heads/master/bridges_yearbuilt.csv


```{r}
url <- "https://raw.githubusercontent.com/acatlin/data/refs/heads/master/bridges_yearbuilt.csv"

df <- read_csv(
  file = url,
  show_col_types = FALSE,
  progress = FALSE
)

glimpse(df)

df |>
  gt()

```

